Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-03 paper/paper.qmd
committing new file onto git for paper.qmd

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(usmap)
library(arrow)
library(knitr)
library(Matrix)
library(splines)
library(kableExtra)
library(broom.mixed)
library(brms)
```

@@

To give an overview of the data, we provide summary statistics of the key variables, highlighting the characteristics of the polling support for each candidate across the battleground states.
```{r}
#| label: tbl-outcomevar
#| fig-cap: "Summary Statistics for Trump and Harris Polling Data. This table presents key summary statistics, including total polls, average support percentages, and the maximum and minimum polling levels for both candidates across the specified battleground states. All values are rounded to three decimal places, providing an overview of the polling landscape as of the selected election date."
#| echo: false
#| warning: false
#| message: false

# Load cleaned data
clean_poll_data <- read_parquet("/Users/jamielee/2024_US_Elections/data/02-analysis_data/analysis_data.parquet")

@@

### Outcome Variable

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stateaverage
#| fig-cap: "Table displaying the percentage of support for Donald Trump and Kamala Harris across key battleground states in the lead-up to the 2024 United States Presidential Election. The table lists states with their respective support percentages, highlighting the competitive landscape as voters prepare to make their decisions on Election Day. Notably, Trump maintains a slight lead in states such as Arizona and Georgia, while Harris shows stronger support in states like Michigan and Nevada. This data provides insight into the current polling dynamics and potential electoral outcomes in these critical states."

### Predictive Var
The predictive variable in our model are 'time' and 'state'

## Other Variables

'numeric grade', 'transparency score', 'pollster'

@@

@@

# Additional Data Details
Given the nature of the presidential election, the electoral college is more impactful and influential on the possible outcome compared to the nationwide popular votes. Therefore, the majority of the pollsters choose to focus on the battleground states to perform their polling, to gain more insight into the public sentiment specific to certain states. For instance, Sienna/NYT mentions that they have in-state interviewers who call their respondents in states such as Arizona, Georgia, Michigan, Nevada, Pennsylvania, and Wisconsin.

>>>>

library(tidyverse)
library(janitor)
library(lubridate)
library(modelsummary)
library(rstanarm)
library(usmap)
library(arrow)
library(knitr)
library(Matrix)
library(splines)
library(kableExtra)
library(broom.mixed)
library(brms)

# Load cleaned data
clean_poll_data <- read_parquet("/Users/jamielee/2024_US_Elections/data/02-analysis_data/analysis_data.parquet")
```

@@

@@

### Outcome Variable
Outcome variable is 'pct'

To give an overview of the outcome variable, divided per candidate, we provide summary statistics of this key variable, highlighting the characteristics of the polling support for each candidate across the battleground states.
```{r}
#| label: tbl-outcomevar
#| fig-cap: "Summary Statistics for Trump and Harris Polling Data. This table presents key summary statistics, including total polls, average support percentages, and the maximum and minimum polling levels for both candidates across the specified battleground states. All values are rounded to three decimal places, providing an overview of the polling landscape as of the selected election date."
#| echo: false
#| warning: false
#| message: false

### Predictive Variables
The predictive variables is 'state' 
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Bar plot showing the distribution of observations across key battleground states in the 2024 United States Presidential Election polling data. The plot illustrates the count of polling observations for each state, highlighting states with more frequent polling activity. This visualization aids in understanding the level of engagement and data collection in each battleground state leading up to the election."

# Visualize the distribution of the state variable
ggplot(clean_poll_data, aes(x = state)) +
  geom_bar(fill = "steelblue") +  # Bar plot with specified fill color
  labs(
    title = "Distribution of State Variable in Polling Data",
    x = "State",
    y = "Count of Observations"
  ) +
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    plot.title = element_text(hjust = 0.5)  # Center the plot title
  )
```

Another predictive variable is 'end-date'
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling end dates for key battleground states in the lead-up to the 2024 United States Presidential Election. The histogram shows the frequency of polls conducted over time, providing insight into the timing of polling activities and potential trends in voter sentiment as Election Day approaches. This visualization is crucial for understanding how polling data correlates with the election timeline."

# Visualize the distribution of end_date
ggplot(clean_poll_data, aes(x = end_date)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +  # Histogram with specified binwidth
  labs(
    title = "Distribution of Polling End Dates",
    x = "End Date",
    y = "Count of Polls"
  ) +
  scale_x_date(date_labels = "%b %d, %Y", date_breaks = "1 week") +  # Format date labels and breaks
  theme_minimal() +  # Use a minimal theme
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Rotate x-axis labels for better readability
    plot.title = element_text(hjust = 0.5)  # Center the plot title
  )

###Other Variables

'numeric grade'
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Histogram illustrating the distribution of polling end dates for key battleground states in the lead-up to the 2024 United States Presidential Election. Displays the frequency of polls conducted over time, providing insight into the timing of polling activities and potential trends in voter sentiment as Election Day approaches. This visualization is crucial for understanding how polling data correlates with the election timeline."

# Density plot for numeric grade
ggplot(clean_poll_data, aes(x = numeric_grade)) +
  geom_density(fill = "lightgreen", alpha = 0.5) +
  labs(
       x = "Numeric Grade",
       y = "Density") +
  theme_minimal()

'transparency score'
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram illustrating the distribution of transparency scores for polls in key battleground states. Reveals the frequency of different transparency ratings, providing insight into the reliability of polling data. Understanding the distribution of transparency scores is crucial for evaluating the credibility of the polling results as Election Day approaches."

# Distribution of transparency score
ggplot(clean_poll_data, aes(x = transparency_score)) +
  geom_histogram(binwidth = 1, fill = "orange", color = "black") +
  labs(
       x = "Transparency Score",
       y = "Count") +
  theme_minimal()

'pollster'
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar graph illustrating the distribution of polls by various pollsters in key battleground states. Displays the count of polls conducted by each pollster, highlighting the most frequently utilized polling organizations. Understanding the diversity of pollsters is essential for assessing the reliability and variance in polling data as Election Day approaches."

# Horizontal bar graph for the distribution of pollsters
ggplot(clean_poll_data, aes(x = factor(pollster))) +  # Convert to factor to ensure proper categorization
  geom_bar(fill = "purple") +
  labs(
       x = "Count of Polls",
       y = "Pollster") +  # Notice the labels are switched for horizontal orientation
  theme_minimal() +
  coord_flip() +  # Flip the coordinates for a horizontal bar graph
  theme(axis.text.y = element_text(angle = 0, hjust = 1))  # Adjust y-axis text if necessary
```

'start_date'
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states in the lead-up to the 2024 United States Presidential Election. This visualization displays the frequency of polls conducted over time, providing insights into when polling activities were initiated and potential trends in voter sentiment as Election Day approaches. Understanding the timing of polls is crucial for evaluating the relevance of the polling data."

# Distribution of start_date with x-axis labels every 15 days
ggplot(clean_poll_data, aes(x = start_date)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black") +
  labs(
       x = "Start Date",
       y = "Count") +
  scale_x_date(date_breaks = "15 days", date_labels = "%b %d") +  # Adjust x-axis labels
  theme_minimal()
```

@@

## State-Level Polling Averages for Trump and Harris
WRITE - explain why we are showing this average table first and what the table reveals
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stateaverage
#| fig-cap: "Table displaying the percentage of support for Donald Trump and Kamala Harris across key battleground states in the lead-up to the 2024 United States Presidential Election. The table lists states with their respective support percentages, highlighting the competitive landscape as voters prepare to make their decisions on Election Day. Notably, Trump maintains a slight lead in states such as Arizona and Georgia, while Harris shows stronger support in states like Michigan and Nevada. This data provides insight into the current polling dynamics and potential electoral outcomes in these critical states."

@@

# Additional Data Details

Given the nature of the presidential election, the electoral college is more impactful and influential on the possible outcome compared to the nationwide popular votes. Therefore, the majority of the pollsters choose to focus on the battleground states to perform their polling, to gain more insight into the public sentiment specific to certain states. For instance, Sienna/NYT mentions that they have in-state interviewers who call their respondents in states such as Arizona, Georgia, Michigan, Nevada, Pennsylvania, and Wisconsin.
